Supplementary Materials

### 2021-10-08 analysis/paper/paper.tex
incorporate AGH suggestions

\usepackage{lineno}
\linenumbers
\ifluatex
  \usepackage{selnolig}  % disable illegal ligatures
\fi
\newlength{\cslhangindent}
\setlength{\cslhangindent}{1.5em}
\newlength{\csllabelwidth}
\setlength{\csllabelwidth}{3em}
\newenvironment{CSLReferences}[2] % #1 hanging-ident, #2 entry spacing
 {% don't indent paragraphs
  \setlength{\parindent}{0pt}
  % turn on hanging indent if param 1 is 1
  \ifodd #1 \everypar{\setlength{\hangindent}{\cslhangindent}}\ignorespaces\fi
  % set entry spacing
  \ifnum #2 > 0
  \setlength{\parskip}{#2\baselineskip}
  \fi
 }%
 {}
\usepackage{calc}
\newcommand{\CSLBlock}[1]{#1\hfill\break}
\newcommand{\CSLLeftMargin}[1]{\parbox[t]{\csllabelwidth}{#1}}
\newcommand{\CSLRightInline}[1]{\parbox[t]{\linewidth - \csllabelwidth}{#1}\break}
\newcommand{\CSLIndent}[1]{\hspace{\cslhangindent}#1}

\title{Investigating Biases Associated with Dietary Starch Incorporation and Retention with an Oral Biofilm Model}
\author{Bjørn Peare Bartholdy \and Amanda G. Henry}
\date{2021-09-30}

\begin{document}
\maketitle
\begin{abstract}
Dental calculus, mineralised plaque, has proven to contain a wealth of information
on the dietary habits of past populations. The extraction and identification of
starch granules contained within dental calculus has already provided valuable
dietary insights from a wide range of regions and time periods, and seem to preserve
well in the archaeological record.
While the majority of studies have focused on microfossil extraction and
identification from archaeological remains, few studies have
addressed the biases of starch retention in dental calculus, and have been
limited to \emph{in vivo} studies on modern humans and non-human primates.
Here, we present a multispecies oral biofilm model, which allows experimental
research to be conducted on \emph{in vitro} dental calculus in a controlled laboratory
setting. The biofilms were exposed treatment solutions with known quantities of
dietary starches (wheat and potato) during the 25-day growth period. After this,
the starch granules were extracted from the mature biofilm (dissolution in EDTA),
and counted. We
show that the granule counts extracted from the model dental calculus represented
a low proportion of the total counts (ranging from 0.064\% to 0.161\%) exposed to
the biofilms throughout the experiment. Additionally, we found that the ratios
of granule sizes from the extracted starch granules differed from the original
treatment solutions, with large granules (\textgreater20 \(\mu\)m) consistently being
under-represented.
Additionally, the size ratios of granules within starch species shifted between
the input starch solutions and the granules extracted from the model calculus,
with large granules (\textgreater20 \(\mu\)m) consistently being underrepresented.
We also found a correlation
between the absolute granule counts and dry-weight of the biofilm
(\emph{r} = 0.66, 90\%CI{[}0.46,0.79{]}), as well
as between the concentration (count per mg) of granules and dry-weight
(\emph{r} = 0.3, 90\%CI{[}0.06,0.51{]}).\\
Our results reinforce previous \emph{in vivo} studies suggesting that dental calculus
presents a very small, and partly biased picture of the original dietary intake
of starches, with an over-representation of plants producing granules smaller than
20 \(\mu\) in size. The experimental model presented here is well-suited to address
the need for further validation of methods and biases associated with dietary
research on dental calculus.
\end{abstract}

@@

\setstretch{2.0}
Word count: 4634

@@

Dental calculus, mineralised dental plaque, has proven to contain a wealth of
dietary information in the form of plant microfossils
(Hardy et al., 2009; Henry \& Piperno, 2008),
proteins (Hendy et al., 2018; Warinner, Hendy, et al., 2014),
and other organic residues (Buckley et al., 2014).
This dietary information can be preserved within calculus over many millenia,
providing a unique window into the food-related behaviours of past populations
(Henry \& Piperno, 2008; Jovanović et al., 2021; Tao et al., 2020)
and extinct species (Hardy et al., 2012; Henry et al., 2014).

Until recently, only a few studies directly investigated the presence of plant
microremains
in the dental calculus of archaeological remains. The ability to extract phytoliths
from the dental calculus of archaeological fauna to investigate diet was first
noted by Armitage (1975),
and later by Middleton and Rovner (1994),
and Fox and colleagues (1996). Starches and phytoliths were
extracted from human dental calculus by Cummings and Magennis (1997).\\
In more recent years, the study of dental calculus has increased exponentially,
and the wealth of information contained within the mineralised matrix has been,
to a larger extent, acknowledged. The applications of dental calculus span a wide
variety of archaeological research areas, such as oral microbiome
characterisation (including pathogens) through the analysis of DNA and proteins
(Adler et al., 2013; Warinner, Rodrigues, et al., 2014),
microbotanical remains (Hardy et al., 2009; Henry \& Piperno, 2008; Mickleburgh \& Pagán-Jiménez, 2012),
other organic residues and proteins from dietary compounds
(Buckley et al., 2014; Hendy et al., 2018),
and nicotine-use (Eerkens et al., 2018). Especially the extraction
of starch granules has become a rich source of dietary
information, as starch granules have proven to preserve well within dental calculus
over a variety of geographical and temporal ranges
(Henry et al., 2014; Jovanović et al., 2021; Piperno \& Dillehay, 2008; Tao et al., 2020).

Despite this, our knowledge of dental calculus and the incorporation pathways of
the various markers is limited (Radini et al., 2017), as is our
knowledge of information-loss caused by these pathways. Additionally, the methods
we use to extract and analyse dental calculus, and make inferences on past diets
represent another potential source of bias. Studies on both archaeological and
modern individuals have explored these biases in more detail.
Extraction methods were tested
by Tromp and colleagues (2017), specifically regarding
decalcification using HCl or EDTA.
The authors found significantly more starches with the EDTA extraction method
than the HCl extraction method; however, as noted by the authors, comparisons
involving archaeological calculus are problematic due to variability between and
within individuals.
Studies conducted on modern humans (Leonard et al., 2015)
and non-human primates (Power et al., 2015),
have explored the how well microremains (phytoliths and starches)
extracted from dental calculus represent the actual dietary intake.
These studies are justifiably limited,
despite meticulous documentation and observation, due to unknown variables and
uncertainty involved when studying living organisms. Dental calculus is a complex
oral biofilm with a multifactorial aetiology and variable formation rates both
within and between individuals (Haffajee et al., 2009; Jepsen et al., 2011), contributing to
the stochasticity of starch representation being observed in numerous studies.
Additionally, the concentration of oral \(\alpha\)-amylase differs both between and
within individuals (Froehlich et al., 1987; Nater et al., 2005),
causing different rates of hydrolysis of the starch granules present in the oral
cavity. Add to this the effects of the many different methods
of starch processing, as well as post-depositional processes that are still being
explored (García-Granero, 2020), and you have a highly unpredictable process.

In this exploratory study, we use an oral biofilm model to investigate the
retention of starch granules within dental calculus in a controlled laboratory
setting, allowing us full control over dietary input. Our main questions concern
the representation of granules extracted
from the calculus compared to the actual intake. How much of the original diet is
incorporated into the calculus, and how much is recovered?
Is there differential loss of information from specific dietary markers that affects
the obtained dietary information, and how does this affect the representation of
diet from extracted microremains?\\
We find that, despite the absence of \(\alpha\)-amylase in
the model, a limited proportion of the starch input is actually
retained in the calculus, as well as a shift in the size ratios of individual
starch granules
that are incorporated into the calculus. We also show that the number of starch
granules that are incorporated increases as the size of the calculus deposit
increases.

@@

In this study we employ a multispecies oral biofilm model following a modified
protocol from Sissons and colleagues (1991) and
Shellis (1978). The setup comprises a polypropylene
24 deepwell PCR plate (KingFisher 97003510) with a lid containing 24 pegs (substrata), which
is autoclaved at 120°C, 1 bar overpressure, for 20 mins.

@@

After 15 days, mineralisation was encouraged with a
calcium phosphate monofluorophosphate urea (CPMU) solution containing
20 mmol/l CaCl\textsubscript{2}, 12 mmol/l NaH\textsubscript{2}PO\textsubscript{4}, 5 mmol/l Na\textsubscript{2}PO\textsubscript{3}F, 500 mmol/l Urea,
and (0.04 g/l MgCl)
(Pearce \& Sissons, 1987; Sissons et al., 1991).
The substrata were submerged in 1 ml/well CPMU for six minutes, five times
daily, in a two-hour cycle. During the mineralisation period, starch treatments
were reduced to once per day after the five CPMU treatments. This process was repeated
for 10 days until the end of the experiment on day 24
(see \ref{fig:protocol-fig} for an overview of the protocol). More detailed protocols
are available on protocols.io.

@@

An \(\alpha\)-amylase (EC 3.2.1.1) activity assay was conducted on artificial
saliva samples collected from the plate wells on days 3, 6, 8, 9, 10, 12, and 14.
Whole saliva samples were collected on days 0, 3, and 5 as positive controls.
Collected samples were stored at 4°C until the assay was conducted on day 18.
All samples and standard curves were run in triplicates on two separate plates.
Positive control saliva samples were compared against a standard curve containing
H\textsubscript{2}O, while artificial saliva samples were compared against a standard curve
containing sterile artificial saliva (due to the colour of artificial saliva).
Two photometric readings were conducted for each plate with a 540 nm filter on a
Multiskan FC Microplate Photometer (Thermo Scientific 51119000).
The protocol is a slightly modified version of an Enzymatic Assay of \(\alpha\)-Amylase
(\url{https://www.sigmaaldrich.com/NL/en/technical-documents/protocol/protein-biology/enzyme-activity-assays/enzymatic-assay-of-a-amylase}) (Bernfeld, 1955), which measures the amount of
maltose released from starch by \(\alpha\)-amylase activity. Results are reported
in units (U) per mL enzyme, where 1 U releases 1 mg of maltose.
For a more detailed protocol, see

@@

All scripts and data used in the analysis are available on OSF
(\url{https://osf.io/uc5qy/}) and Github (\url{https://github.com/bbartholdy/byoc-starch}).
All protocols are available on protocols.io.

@@

\begin{figure}
\centering
\includegraphics{/home/bjorn/Documents/Uni/publications/PhD/byocstarch/analysis/figures/starches_w_bar.jpg}
\caption{\label{fig:microscope-fig}Microscope image of potato starches on a microscope slide, as well as a bacterial community. Scale bar = 20 \(\mu\)m.}
\end{figure}

@@

\caption{\label{tab:sample-prop-tbl}The mean percentage of starches from the solutions that 
              were incorported into the samples.}
\centering
\begin{tabular}[t]{l|l|l|l|l|l}
\hline
Treatment & Starch & Small & Medium & Large & Total\\
\hline
mix & potato &  & 0.186\% & 0.054\% & 0.124\%\\
\hline
mix & wheat & 0.051\% & 0.102\% & 0.077\% & 0.064\%\\
\hline
mix & both & 0.051\% & 0.114\% & 0.069\% & 0.069\%\\
\hline
potato & potato & 0.284\% & 0.266\% & 0.060\% & 0.161\%\\
\hline
wheat & wheat & 0.094\% & 0.189\% & 0.069\% & 0.109\%\\
\hline
\end{tabular}
\end{table}

The proportion of total starches from the solutions that were incorporated
into the samples ranged from
0.064\% to 0.161\%,
with potato granules being more readily incorporated than wheat in both the
separated- and mixed-treatment samples
(Table \ref{tab:sample-prop-tbl}).
There is an inverse relationship between the absolute starch count in the solutions
and the proportional incorporation of starches in the biofilm samples, i.e.~potato
had the lowest absolute count in solutions, but the highest proportional
incorporation, and vice versa for the mixed treatment.

@@

\begin{figure}
\centering
\includegraphics{../figures/cor-plot-1.pdf}
\caption{\label{fig:cor-plot}Scatter plot of sample weight and standardised starch count by Z-score for seprated treatments.}
\end{figure}

@@

Here, we have provided a method for exploring the incorporation of dietary
starches into the mineral matrix of a dental calculus biofilm model. Our results show
that a very low proportion of the starches exposed to the biofilm during growth are
retained in the mineral matrix, and that the size of the starch granules
may affect the likelihood of incorporation. The proportions of starch granules
(of all sizes) present in the extracted samples were similar across all treatments
(0.064\% to 0.161\%),
despite large differences in absolute granule counts between wheat
(mean = 25,404,000)
and potato
(mean = 3,016,000)
solutions.\\
The absolute counts, however, differed more visibly between treatments and was
proportional with the total count of granules in the treatment solutions. Wheat
and mixed solutions had the highest absolute mean count of starch granules, and
also had the highest absolute mean count of starch granules extracted from the
dental calculus
(Tables \ref{tab:solution-count-tbl} and \ref{tab:sample-count-tbl}).
This suggests that the starches that are more frequently consumed will be present
in higher quantities in the dental calculus, at least prior to inhumation and
degradation in the burial environment.
Despite the low proportion of granules recovered from the model calculus
(0.064\% to 0.161\%),
the absolute counts were still substantially greater than counts recovered from
archaeological remains
(Tromp et al., 2017; Tromp \& Dudgeon, 2015; Wesolowski et al., 2010). This could, in part, be due to the
activity of oral amylase, which was absent in our model.

>>>>

\usepackage{lineno}
\linenumbers
\usepackage{subfig}
\ifluatex
  \usepackage{selnolig}  % disable illegal ligatures
\fi
\newlength{\cslhangindent}
\setlength{\cslhangindent}{1.5em}
\newlength{\csllabelwidth}
\setlength{\csllabelwidth}{3em}
\newenvironment{CSLReferences}[2] % #1 hanging-ident, #2 entry spacing
 {% don't indent paragraphs
  \setlength{\parindent}{0pt}
  % turn on hanging indent if param 1 is 1
  \ifodd #1 \everypar{\setlength{\hangindent}{\cslhangindent}}\ignorespaces\fi
  % set entry spacing
  \ifnum #2 > 0
  \setlength{\parskip}{#2\baselineskip}
  \fi
 }%
 {}
\usepackage{calc}
\newcommand{\CSLBlock}[1]{#1\hfill\break}
\newcommand{\CSLLeftMargin}[1]{\parbox[t]{\csllabelwidth}{#1}}
\newcommand{\CSLRightInline}[1]{\parbox[t]{\linewidth - \csllabelwidth}{#1}\break}
\newcommand{\CSLIndent}[1]{\hspace{\cslhangindent}#1}

\title{Investigating Biases Associated with Dietary Starch Incorporation and Retention with an Oral Biofilm Model}
\author{Bjørn Peare Bartholdy \and Amanda G. Henry}
\date{2021-10-08}

\begin{document}
\maketitle
\begin{abstract}
Dental calculus has proven to contain a wealth of information
on the dietary habits of past populations. These insights have, to a large extent,
been obtained by the extraction and identification of
starch granules contained within the mineralised dental plaque from a wide range
of regions and time periods. The scope of previous studies have been limited to
microfossil extraction and identification to reconstruct dietary preferences from
the archaeological record, and few studies have attempted to address the biases
of starch retention in dental calculus. Those that have considered this problem
have been limited to \emph{in vivo} studies on modern humans and non-human primates.
Here, we present a multispecies oral biofilm model, which allows experimental
research on starch incorporation and retention to be conducted on \emph{in vitro}
dental calculus in a controlled laboratory setting. The biofilms were exposed
to treatment solutions with known quantities of
dietary starches (wheat and potato) during the 25-day growth period. After this,
the starch granules were extracted from the mature biofilm (by dissolution in EDTA),
and counted. We show that the granule counts extracted from the model dental calculus
represented a low proportion (ranging from 0.064\% to 0.161\%) of the total number
of granules exposed to the biofilms throughout the experiment.
Additionally, we found that the ratios of granule sizes from the extracted
starch granules differed from the original treatment solutions, with large
granules (\textgreater20 \(\mu\)m) consistently being under-represented.
We also found a correlation
between the absolute granule counts and dry-weight of the biofilm
(\emph{r} = 0.66, 90\%CI{[}0.46,0.79{]}), as well
as between the concentration (count per mg) of granules and dry-weight
(\emph{r} = 0.30, 90\%CI{[}0.06,0.51{]}).\\
Our results reinforce previous \emph{in vivo} studies suggesting that dental calculus
presents a very small, and partly biased picture of the original dietary intake
of starches, with an over-representation of plants producing granules smaller than
20 \(\mu\)m in size. The experimental model presented here is well-suited to address
the need for further validation of methods and biases associated with dietary
research on dental calculus.
\end{abstract}

@@

\setstretch{2.0}
Word count: 4682

@@

Dental calculus has proven to contain a wealth of
dietary information in the form of plant microfossils
(Hardy et al., 2009; Henry \& Piperno, 2008),
proteins (Hendy et al., 2018; Warinner, Hendy, et al., 2014),
and other organic residues (Buckley et al., 2014).
This dietary information can be preserved within the mineralised dental plaque
over many millennia, providing a unique window into the food-related behaviours of
past populations
(Henry \& Piperno, 2008; Jovanović et al., 2021; Tao et al., 2020)
and extinct species (Hardy et al., 2012; Henry et al., 2014).

Until recently, only a few studies directly investigated the presence of plant
microremains
in the dental calculus of archaeological remains. The ability to extract phytoliths
from the dental calculus of archaeological fauna to investigate diet was first
noted by Armitage (1975),
and later by Middleton and Rovner (1994),
and Fox and colleagues (1996). Starches and phytoliths were
extracted from human dental calculus by Cummings and Magennis (1997).\\
In more recent years, the study of dental calculus has increased exponentially,
and the wealth of information contained within the mineralised matrix has largely
been acknowledged. The use of dental calculus spans a wide
variety of archaeological research areas, such as oral microbiome
characterisation (including pathogens) through the analysis of DNA and proteins
(Adler et al., 2013; Warinner, Rodrigues, et al., 2014),
microbotanical remains (Hardy et al., 2009; Henry \& Piperno, 2008; Mickleburgh \& Pagán-Jiménez, 2012),
other organic residues and proteins from dietary compounds
(Buckley et al., 2014; Hendy et al., 2018),
and nicotine-use (Eerkens et al., 2018). Especially the extraction
of starch granules has become a rich source of dietary
information, as starch granules have proven to preserve well within dental calculus
over a variety of geographical and temporal ranges
(Henry et al., 2014; Jovanović et al., 2021; Piperno \& Dillehay, 2008; Tao et al., 2020).

Despite this, our knowledge of dental calculus and the incorporation pathways of
the various markers is limited (Radini et al., 2017), as is our
knowledge of information-loss caused by these pathways. Additionally, the methods
we use to extract and analyse dental calculus, and make inferences on past diets
represent another potential source of bias. Studies on both archaeological and
modern individuals have explored these biases in more detail.
Extraction methods were tested
by Tromp and colleagues (2017), specifically regarding
decalcification using HCl or EDTA.
The authors found significantly more starches with the EDTA extraction method
than the HCl extraction method; however, as noted by the authors, comparisons
involving archaeological calculus are problematic due to variability between and
within individuals.
Studies conducted on modern humans (Leonard et al., 2015)
and non-human primates (Power et al., 2015),
have explored the how well microremains (phytoliths and starches)
extracted from dental calculus represent the actual dietary intake.
These studies are justifiably limited,
despite meticulous documentation and observation, due to unknown variables and
uncertainty involved when studying living organisms. Dental calculus is a complex
oral biofilm with a multifactorial aetiology and variable formation rates both
within and between individuals (Haffajee et al., 2009; Jepsen et al., 2011), contributing to
the stochasticity of starch representation being observed in numerous studies.
Additionally, the concentration of oral \(\alpha\)-amylase differs both between and
within individuals (Froehlich et al., 1987; Nater et al., 2005),
causing different rates of hydrolysis of the starch granules present in the oral
cavity. Add to this the effects of the many different methods
of starch processing, as well as post-depositional processes that are still being
explored (García-Granero, 2020), and it becomes clear that using
dental calculus to reconstruct diet is a highly unpredictable process.

In this exploratory study, we use an oral biofilm model to investigate the
retention of starch granules within dental calculus in a controlled laboratory
setting, allowing us full control over dietary input. Our main questions concern
the representation of granules extracted
from the calculus compared to the actual intake. How much of the original diet is
incorporated into the calculus, and how much is recovered?
Is there differential loss of information from specific dietary markers that affects
the obtained dietary information, and how does this affect the representation of
diet from extracted microremains?\\
We find that, despite the absence of \(\alpha\)-amylase in
the model, a limited proportion of the starch input is actually
retained in the calculus. We also observed a shift in the size ratios of individual
starch granules that are incorporated into the calculus, and that the number of
incorporated starch granules increases as the size of the calculus deposit
increases.

@@

In this study we employ a multispecies oral biofilm model following a modified
protocol from Sissons and colleagues (1991) and
Shellis (1978). In brief, a biofilm inoculated
with whole saliva was grown on a substrate suspended in artificial saliva, and
fed with sugar (sucrose). After several days of growth, the biofilm was exposed
to starch solutions. Mineralisation of the biofilm was aided by exposure to a
calcium phosphate solution. After 25 days of growth, the mineralised biofilm was
collected for further analysis. The setup comprises a polypropylene
24 deepwell PCR plate (KingFisher 97003510) with a lid containing 24 pegs (substrata), which
is autoclaved at 120°C, 1 bar overpressure, for 20 mins.

@@

After 15 days, mineralisation was encouraged with a
calcium phosphate monofluorophosphate urea (CPMU) solution containing
20 mmol/l CaCl\textsubscript{2}, 12 mmol/l NaH\textsubscript{2}PO\textsubscript{4}, 5 mmol/l Na\textsubscript{2}PO\textsubscript{3}F, 500 mmol/l Urea,
and (0.04 g/l MgCl)
(Pearce \& Sissons, 1987; Sissons et al., 1991).
The substrata were submerged in 1 ml/well CPMU for six minutes, five times
daily, in a two-hour cycle. During the mineralisation period, starch treatments
were reduced to once per day after the five CPMU treatments. This process was repeated
for 10 days until the end of the experiment on day 24
(see \ref{fig:protocol-fig} for an overview of the protocol).

@@

An \(\alpha\)-amylase (EC 3.2.1.1) activity assay was conducted on artificial
saliva samples collected from the plate wells on days 3, 6, 8, 9, 10, 12, and 14.
Whole saliva samples were collected on days 0, 3, and 5 as positive controls.
Collected samples were stored at 4°C until the assay was conducted on day 18.
All samples and standard curves were run in triplicates on two separate plates.
Positive control saliva samples were compared against a standard curve containing
H\textsubscript{2}O, while artificial saliva samples were compared against a standard curve
containing sterile artificial saliva (due to the colour of artificial saliva).
Two photometric readings were conducted for each plate with a 540 nm filter on a
Multiskan FC Microplate Photometer (Thermo Scientific 51119000).
The protocol is a slightly modified version of an Enzymatic Assay of \(\alpha\)-Amylase
(\url{https://www.sigmaaldrich.com/NL/en/technical-documents/protocol/protein-biology/enzyme-activity-assays/enzymatic-assay-of-a-amylase}) (Bernfeld, 1955), which measures the amount of
maltose released from starch by \(\alpha\)-amylase activity. Results are reported
in units (U) per mL enzyme, where 1 U releases 1 mg of maltose.

@@

All scripts and data used in the analysis are available on OSF
(\url{https://osf.io/uc5qy/}) and Github (\url{https://github.com/bbartholdy/byoc-starch}).
More detailed protocols are available on protocols.io.

@@

\begin{figure}
\includegraphics[width=0.5\linewidth]{/home/bjorn/Documents/Uni/publications/PhD/byocstarch/analysis/figures/starches_w_bar} \caption{Microscope image of a biofilm sample exposed to the potato starch solution. Potato granules can be seen within a bacterial community. Scale bar = 20 $\mu$m.}\label{fig:microscope-fig}
\end{figure}

@@

\caption{\label{tab:sample-prop-tbl}The mean percentage of starches from the solutions that 
              were incorporated into the samples.}
\centering
\begin{tabular}[t]{l|l|l|l|l|l}
\hline
Treatment & Starch & Small & Medium & Large & Total\\
\hline
mix & potato &  & 0.186\% & 0.054\% & 0.124\%\\
\hline
mix & wheat & 0.051\% & 0.102\% & 0.077\% & 0.064\%\\
\hline
mix & both & 0.051\% & 0.114\% & 0.069\% & 0.069\%\\
\hline
potato & potato & 0.284\% & 0.266\% & 0.060\% & 0.161\%\\
\hline
wheat & wheat & 0.094\% & 0.189\% & 0.069\% & 0.109\%\\
\hline
\end{tabular}
\end{table}

The proportion of total starches from the solutions that were incorporated
into the samples ranged from
0.064\% to 0.161\%,
with potato granules being more readily incorporated than wheat in both the
separated- and mixed-treatment samples
(Table \ref{tab:sample-prop-tbl}).
There is an inverse relationship between the absolute starch count in the solutions
and the proportional incorporation of starches in the biofilm samples, i.e., potato
had the lowest absolute count in solutions, but the highest proportional
incorporation, and vice versa for the mixed treatment.

@@

\begin{figure}
\centering
\includegraphics{../figures/cor-plot-1.pdf}
\caption{\label{fig:cor-plot}Scatter plot of sample weight and standardised starch count by Z-score for separated treatments.}
\end{figure}

@@

Here, we have provided a method for exploring the incorporation of dietary
starches into the mineral matrix of a dental calculus biofilm model. Our results show
that a very low proportion of the starches exposed to the biofilm during growth are
retained in the mineral matrix, and that the size of the starch granules
may affect the likelihood of incorporation. The proportions of starch granules
(of all sizes) present in the extracted samples were similar across all treatments
(0.064\% to 0.161\%),
despite large differences in absolute granule counts between wheat
(mean = 25,404,000)
and potato
(mean = 3,016,000)
solutions.\\
The absolute counts, however, differed more visibly between treatments and was
proportional with the total count of granules in the treatment solutions. Wheat
and mixed solutions had the highest absolute mean count of starch granules, and
also had the highest absolute mean count of starch granules extracted from the
dental calculus
(Tables \ref{tab:solution-count-tbl} and \ref{tab:sample-count-tbl}).
This suggests that the starches that are more frequently consumed will be present
in higher quantities in the dental calculus, at least prior to inhumation and
degradation in the burial environment.
Despite the low proportion of granules recovered from the model calculus
(0.064\% to 0.161\%),
the absolute counts were still substantially greater than counts recovered from
archaeological remains
(Tromp et al., 2017; Tromp \& Dudgeon, 2015; Wesolowski et al., 2010), which could in part be due to the
lack of oral amylase activity in our model.

### 2021-10-08 analysis/supplementary-materials/supp-mat.Rmd
incorporate AGH suggestions

@@

```{r img-1, echo=FALSE, fig.cap="Image of starch granules extracted from a calculus sample"}

```

>>>>

## Experimental setup

```{r plate-lid-1, fig.cap="The 24 deepwell plate and the lid with pegs (substrata)"}
knitr::include_graphics(here("analysis/figures/plate_lid_side.jpg"))
```

```{r plate-lid-2, fig.cap="The 24 deepwell plate with the lid on (sort of...)"}
knitr::include_graphics(here("analysis/figures/plate_lid_on.jpg"))
```

@@

```{r img-1, echo=FALSE, fig.cap="Image of starch granules extracted from a potato treatment sample"}
knitr::include_graphics(here("analysis/figures/starches_w_bar.jpg"))
```

```{r img-2}
knitr::include_graphics(here("analysis/figures/SNAP-103412-0006.jpg"))
```

```{r img-3, fig.cap="Microscope image of wheat starch from a wheat treatment sample."}
knitr::include_graphics(here("analysis/figures/SNAP-164650-0012.jpg"))
```
